### 2024-06-05 thesis/appendix.tex
cleaned all problems and warnings
final thesis

\subsection{Cross validation for time series data}
When data points are dependent on preceding values, we cannot use standard $K$-fold cross validation. The rationale is that $K$-fold will randomise the order of the data, thus it might happen to use future data to predict the past; we want to avoid such behaviour in a time series setting. 
When carrying out out any kind of cross validation, we must keep consistency in the way we evaluate our predictors during model selection and in the way we perform evaluation of the test data.
Hence, for time series crossvalidation we have the timeseries split procedure, see Figure \ref{fig:crossvalidationtimeseries} for a visualisation. The blue observations make up the training sets while the orange observations form the test sets.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries.png}
    \caption{Cross validation for one step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries}
\end{figure}
We then average all the observed losses in order to get an estimate for $Err$.
Notice, in the literature this procedure is sometimes also referred to evaluation on a rolling forecasting origin. This comes from the fact that at each iteration we push forward the origin of our forecast.
The same concept applies for multi step ahead forecasting, see Figure \ref{fig:crossvalidationtimeseries2}.
In predicting $\hat{y}_{N+m}$ we use as inputs $y_1, y_2, \dots, \hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$, where $\hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$ are one step ahead forecasts.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries2.png}
    \caption{Cross validation for $m$ step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries2}
\end{figure}

@@

\newpage
\section{Source code}\label{src_code}
The whole code for the project is hosted on
\href{https://github.com/luca-pernigo/ThesisKernelMethods}{github repo}.

\begin{itemize}
    \item query/: folder containing Scopus data and scripts to generate bibliometric survey plots in Section \ref{literature_review}. Database queries have been taken from \cite{nowotarski}.
    \item thesis/: folder containing tex files for thesis pdf.
    \item beamer/: folder containing tex files for beamer pptx %pptx thesis defence.
    \item experiments: folder containing train, test script for each of the experiments carried out in the thesis alongside with the binary files of the pickled models.
    \item experiments/src/kernel\_quantile\_regression/kqr.py: file implementing our custom kernel quantile regression.
    \item experiments/Data: cleaned data used in our experimental studies.
    \item experiment/plots/: folder storing plots resulting from experiments.
    \item experiment/eda/: scripts for exploratory data analysis.
    \item experiments/point/: scripts for experiments point framework.
    \item experiments/train\_test/: scripts for training, testing and generating tables for experiments probabilistic framework.
    \item experiments/utils/: utility functions used in extracting, transforming and cleaning raw data.
    \item extradata/: folder containing data for experiments in section \ref{appendix:quantile_regressor_extensive_comparison}.
    \item requirements.txt: pip freeze of python packages used.
\end{itemize}

>>>>

\subsection{Cross validation for time series data}
When data points are dependent on preceding values, we cannot use standard $K$-fold cross validation. The rationale is that $K$-fold will randomise the order of the data, thus it might happen to use future data to predict the past; we want to avoid such behaviour in a time series setting. 
When carrying out any kind of cross validation, we must keep consistency in the way we evaluate our predictors during model selection and in the way we perform evaluation of the test data.
Hence, for time series crossvalidation we have the timeseries split procedure, see Figure \ref{fig:crossvalidationtimeseries} for a visualisation. The blue observations make up the training sets while the orange observations form the test sets.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries.png}
    \caption{Cross validation for one step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries}
\end{figure}
We then average all the observed losses in order to get an estimate for $Err$.
Notice, in the literature this procedure is sometimes also referred to evaluation on a rolling forecasting origin. This comes from the fact that at each iteration we push forward the origin of our forecast.
The same concept applies for multi step ahead forecasting, see Figure \ref{fig:crossvalidationtimeseries2}.
In predicting $\hat{y}_{N+m}$ we use as inputs $y_1, y_2, \dots, \hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$, where $\hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$ are one step ahead forecasts.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries2.png}
    \caption{Cross validation for $m$ step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries2}
\end{figure}

@@

\newpage
\section{Source code}\label{src_code}
The whole code for the project is hosted on its
\href{https://github.com/luca-pernigo/ThesisKernelMethods}{github repo}.

\begin{itemize}
    \item query/: folder containing Scopus data and scripts to generate bibliometric survey plots in Section \ref{literature_review}. Database queries have been taken from \cite{nowotarski}.
    \item thesis/: folder containing tex files for thesis pdf.
    \item beamer/: folder containing tex files for beamer pptx. %pptx thesis defence.
    \item experiments: folder containing train, test script for each of the experiments carried out in the thesis alongside with the binary files of the pickled models.
    \item experiments/src/kernel\_quantile\_regression/kqr.py: file implementing our custom kernel quantile regression.
    \item experiments/Data: cleaned data used in our experimental studies.
    \item experiment/plots/: folder storing plots resulting from experiments.
    \item experiment/eda/: scripts for exploratory data analysis.
    \item experiments/point/: scripts for experiments point framework.
    \item experiments/train\_test/: scripts for training, testing and generating tables for experiments probabilistic framework.
    \item experiments/utils/: utility functions used in extracting, transforming and cleaning raw data.
    \item extradata/: folder containing data for experiments in section \ref{appendix:quantile_regressor_extensive_comparison}.
    \item requirements.txt: pip freeze of python packages used.
\end{itemize}

### 2024-06-05 thesis/input_files/literaturereview.tex
final thesis

Kaur et al.\ \cite{kaur2022energy} carries out a comparative study of techniques spanning ideas from statistic and artificial intelligence.

@@

%- Kernel history evolution.
Their name comes from the German word kern, which translates to core in english. Such term was first used by David Hilbert in his paper on integral equations \cite{hilbert} where he introduced the term "definite kernels". Following, Hilbert's and Schmidt's work, \cite{schmidt} lead to the introduction of a new space, the Hilbert space.
In 1909, James Mercer improved Hilbert's work by proposing his theorem \cite{mercer}. This theorem underlies the power of kernel methods, that is the kernel trick.
In 1938, Schoenberg \cite{schoenberg} developed the mathematical results that allow us to find the kernel associated to a specific feature space metric.
In 1941, Kolmogorov \cite{kolmogorov} carried out stuides on representing kernels in linear spaces.
In 1950, Aronszajn \cite{aronszajn} published the first work on reproducing kernel Hilbert spaces (RKHS); developing the general method for representing kernels in linear spaces.
In 1964, Aizerman \cite{aizerman} further improved the theory of RKHS.
It was in the 1990s that the theory of kernel methods got popular, particularly in the field of machine learning. Kernels have been used in various different tasks such as support vector machines (SVM) \cite{vapnik1} \cite{vapnik2}, Gaussian process classifiers \cite{williams}, spline methods \cite{wahba}, neural networks \cite{poggio} and principal component analysis \cite{pca_scholkopf}.
Nevertheless kernel methods received very little attention in the specific setting of EF literature.

@@

Kanagawa et Fukumizu introduces to the concept of kernel mean embedding \cite{pmlr}. Muandet et al.\ \cite{Muandet_2017} surveys established results and new advances in the theory of Hilbert space distribution embeddings. 
It has to be said that, computing and storing such embeddings becomes prohibitive for large scale settings. Rudi et al.\ \cite{2022nystrom} proposes an efficient approximation procedure based on the Nyström method \cite{nystrom}, providing also an upper bound for the approximation error.
Another example of the applicability of kernel methods is \cite{supersamples}. In that paper, Smola et al.\ kernelize the herding algorithm introduced in \cite{welling2009herding,welling2009herding,Welling2010}. The original herding algorithm converts empirical moments into a sequence of pseudo-samples approximating the probability density function (PDF) of the observed random variable. Its weakness is that it can only take a finite number of features and this limits its expressive power.
Smola et al.\ overcome this by translating the algorithm into the framework of kernels.

>>>>

Kaur et al.\ \cite{kaur2022energy} carries out a comparative study of techniques spanning ideas from statistics and artificial intelligence.

@@

%- Kernel history evolution.
Their name comes from the German word kern, which translates to core in english. Such term was first used by David Hilbert in his paper on integral equations \cite{hilbert} where he introduced the term "definite kernels". Following, Hilbert's and Schmidt's work, \cite{schmidt} lead to the introduction of a new space, the Hilbert space.
In 1909, James Mercer improved Hilbert's work by proposing his theorem \cite{mercer}. This theorem underlies the power of kernel methods, that is the kernel trick.
In 1938, Schoenberg \cite{schoenberg} developed the mathematical results that allow us to find the kernel associated to a specific feature space metric.
In 1941, Kolmogorov \cite{kolmogorov} carried out stuides on representing kernels in linear spaces.
In 1950, Aronszajn \cite{aronszajn} published the first work on reproducing kernel Hilbert spaces (RKHS); developing the general method for representing kernels in linear spaces.
In 1964, Aizerman \cite{aizerman} further improved the theory of RKHS.
It was in the 1990s that the theory of kernel methods got popular, particularly in the field of machine learning. Kernels have been used in various different tasks such as support vector machines (SVM) \cite{vapnik1, vapnik2}, Gaussian process classifiers \cite{williams}, spline methods \cite{wahba}, neural networks \cite{poggio} and principal component analysis \cite{pca_scholkopf}.
Nevertheless kernel methods received very little attention in the specific setting of EF literature.

@@

Kanagawa et Fukumizu introduces to the concept of kernel mean embedding \cite{pmlr}. Muandet et al.\ \cite{Muandet_2017} surveys established results and new advances in the theory of Hilbert space distribution embeddings. 
It has to be said that, computing and storing such embeddings becomes prohibitive for large scale settings. Rudi et al.\ \cite{2022nystrom} proposes an efficient approximation procedure based on the Nyström method \cite{nystrom}, providing also an upper bound for the approximation error.
Another example of the applicability of kernel methods is \cite{supersamples}. In that paper, Smola et al.\ kernelize the herding algorithm introduced in \cite{welling2009herding,welling2009herding,Welling2010}. The original herding algorithm converts empirical moments into a sequence of pseudo-samples approximating the probability density function (PDF) of the observed random variable. Its weakness is that it can only take a finite number of features and this limits its expressive power.
Smola et al.\ overcomes this by translating the algorithm into the framework of kernels.

### 2024-06-05 thesis/input_files/metrics.tex
final thesis

Proper evaluation methods guide researchers in choosing the model that best fits their needs. This chapter is dedicated to the most common evaluation metrics adopted by academics in the field of electricity forecasting. Error metrics and measures vary depending on whether we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{Mean absolute error}\label{mae}
Consider the time series with actual values given by $y=(y_{n+1}, y_{n+2},\dots, y_{n+h})$
and its $h$ step ahead point forecast $\hat{y}=(\hat{y}_{n+1}, \hat{y}_{n+2},\dots, \hat{y}_{n+h})$ the mean absolute error (MAE) is defined as
\begin{definition}
    The mean absolute error is defined as
    $$
    \mathrm{MAE}(y,\hat{y})=\frac{1}{h}\| y- \hat{y}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|y_{n+k}-\hat{y}_{n+k}|
    $$
\end{definition}

@@

\section{Pinball score}\label{pinball}
The pinball score or quantile score is used to measure the accuracy of a quantile forecast.
\begin{definition}
    The pinball loss is defined as
    $$
    \mathrm{Pinball}(y_{t},\hat{y}_{t,q},q)=
\begin{cases}
(q-1)(\hat{y}_{t,q}-y_{t}) & y_t > \hat{y}_{t,q} \\
q(\hat{y}_{t,q}-y_t) & y_t \leq \hat{y}_{t,q}
\end{cases}
$$
\end{definition}
The pinball loss is an asymmetric function, it weights its score differently depending on the error sign and on the quantile considered, see Figure \ref{fig:pinball}.
\begin{figure}
    \includegraphics[width=\textwidth]{images/pinball_loss.png}
    \caption{Pinball loss}
    \label{fig:pinball}
  \end{figure}
By averaging all the pinball losses over all quantiles and over the whole forecast horizon, we obtain the pinball loss of the probabilistic forecast.
\section{Winkler score}
\begin{definition}
    The Winkler score is defined as
    $$
    \mathrm{Winkler}(y_t,  u_{t},l_{t},\alpha)=\begin{cases}
        \delta & l_{t}\leq y_{t}\leq u_{t}\\
        \delta+2(l_{t}-y_{t})/\alpha & y_{t}< l_{t}\\
        \delta+2(l_{t}-u_{t})/\alpha & y_{t} \geq u_{t}
    \end{cases}
    $$
\end{definition}
Where $\delta$ is the prediction interval (PI) width, that is $\delta=u_t-l_t$, $\alpha$ is (1-coverage level), $u_t$ is the prediction interval upper threshold and $l_t$ is the prediction interval lower threshold at time step $t$. This score penalises observations falling outside the prediction interval and rewards narrow prediction intervals.
\section{Continous ranked probability score}
The continous ranked probability score (CRPS) measures the difference between the estimated cumulative distribution $\hat{F}$ and the empirical cumulative density function (CDF).
\begin{definition}\label{def_crps}
    The continous ranked probability score is defined as
    $$
    \mathrm{CRPS}(y, \hat{F})=\int\limits_{-\infty}^{\infty}\left(\hat{F}(x)-\mathbb{I}_{\{x-y\}} \right)^2 dx
    $$
\end{definition}
% Nevertheless, we can evaluate the integral in closed form. 
Where the indicator function is defined as 
    $$\mathbb{I}_{\{z\}}=
\begin{cases}
0, & z<0\\
1, & z \geq 0
\end{cases}$$
\\
For a visualisation see Figure \ref{fig:crps}. The grey area is what contributes toward the CRPS score.
The better the estimated cumulative density function is the smaller the total CRPS score will be.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crps.png}
    \caption{CRPS integral \cite{haben2023core}}
    \label{fig:crps}
  \end{figure}
It is worth noting, that the CRPS integral can be rewritten in terms of expectations. This makes its evaluation easier, since we know that the sample mean converges to the expectation for sufficiently big sample sizes by the law of large numbers. This was first pointed out by \cite{proper_scores}, where the authors take advantage of Lemma 2.2 of \cite{new_multi_test2} or equivalently identity (17) of \cite{new_multi_tes1}.
\\
\begin{lemma}
    Let $X_1$, $X_2$, $Y_1$, $Y_2$ be independent real random variables with finite expectations. Let $X_1,X_2$ be identically distributed with distribution function $F$ and let $Y_1,Y_2$ be identically distributed with distribution function $G$. Then
   \begin{equation}\label{eq:crps_expectation}
    \mathbb{E}(|X_1-Y_1|)-\frac{1}{2}\mathbb{E}(|X_1-X_2|)-\frac{1}{2}\mathbb{E}(|Y_1-Y_2|)=\int\limits_{-\infty}^{\infty}\left(F(x)-G(x)\right)^2dx
\end{equation}
\end{lemma}
Notice that, in our case, Equation \ref{eq:crps_expectation}, the distribution $G$ of $Y_1$ and $Y_2$ is degenerate, with all probability mass on a single point $y$.
%  (x notation will need to be cleaned here). Since $G(t)=:$
It follows that the third addend in the summation is zero. That is because $Y_1$ and $Y_2$ both following distribution $G$ implies that $\mathbb{E}(|Y_1-Y_2|)$ corresponds to the difference of two equal constant numbers.

>>>>

Proper evaluation methods guide researchers in choosing the model that best fits their needs. This chapter is dedicated to the most common evaluation metrics adopted by academics in the field of electricity forecasting. Error metrics and measures vary depending on whether we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{Mean absolute error}\label{mae}
Consider the time series with actual values given by $y=(y_{n+1}, y_{n+2},\dots, y_{n+h})$
and its $h$ step ahead point forecast $\hat{y}=(\hat{y}_{n+1}, \hat{y}_{n+2},\dots, \hat{y}_{n+h})$
\begin{definition}
    The mean absolute error is defined as
    $$
    \mathrm{MAE}(y,\hat{y})=\frac{1}{h}\| y- \hat{y}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|y_{n+k}-\hat{y}_{n+k}|
    $$
\end{definition}

@@

\section{Pinball score}\label{pinball}
The pinball score or quantile score is used to measure the accuracy of a quantile forecast.
\begin{definition}
    The pinball loss is defined as
    $$
    \mathrm{Pinball}(y_{t},\hat{y}_{t,q},q)=
\begin{cases}
q(y_{t}-\hat{y}_{t,q}) & y_t > \hat{y}_{t,q} \\
-(q-1)(y_{t}-\hat{y}_{t,q}) & y_t \leq \hat{y}_{t,q}
\end{cases}
$$
\end{definition}
The pinball loss is an asymmetric function, it weights its score differently depending on the error sign and on the quantile considered, see Figure \ref{fig:pinball}.
\begin{figure}
    \includegraphics[width=\textwidth]{images/pinball_loss.png}
    \caption{Pinball loss}
    \label{fig:pinball}
  \end{figure}
By averaging all the pinball losses over all quantiles and over the whole forecast horizon, we obtain the pinball loss of the probabilistic forecast.
\section{Winkler score}
\begin{definition}
    The Winkler score is defined as
    $$
    \mathrm{Winkler}(y_t,  u_{t},l_{t},\alpha)=\begin{cases}
        \delta & l_{t}\leq y_{t}\leq u_{t}\\
        \delta+2(l_{t}-y_{t})/\alpha & y_{t}< l_{t}\\
        \delta+2(l_{t}-u_{t})/\alpha & y_{t} \geq u_{t}
    \end{cases}
    $$
\end{definition}
Where $\delta$ is the prediction interval (PI) width, that is $\delta=u_t-l_t$, $\alpha$ is (1-coverage level), $u_t$ is the prediction interval upper threshold and $l_t$ is the prediction interval lower threshold at time step $t$. This score penalises observations falling outside the prediction interval and rewards narrow prediction intervals.
\section{Continous ranked probability score}
The continous ranked probability score (CRPS) measures the difference between the estimated cumulative distribution $\hat{F}$ and the empirical cumulative density function (CDF).
\begin{definition}\label{def_crps}
    The continous ranked probability score is defined as
    $$
    \mathrm{CRPS}(y, \hat{F})=\int\limits_{-\infty}^{\infty}\left(\hat{F}(x)-\mathbb{I}_{\{x-y\}} \right)^2 dx
    $$
\end{definition}
% Nevertheless, we can evaluate the integral in closed form. 
Where the indicator function is defined as 
    $$\mathbb{I}_{\{z\}}=
\begin{cases}
0, & z<0\\
1, & z \geq 0
\end{cases}$$
\\
For a visualisation see Figure \ref{fig:crps}. The grey area is what contributes toward the CRPS score.
The better the estimated cumulative density function is the smaller the total CRPS score will be.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crps.png}
    \caption{CRPS integral \cite{haben2023core}}
    \label{fig:crps}
  \end{figure}
It is worth noting, that the CRPS integral can be rewritten in terms of expectations. This makes its evaluation easier, since we know that the sample mean converges to the expectation for sufficiently big sample sizes by the law of large numbers. This was first pointed out by \cite{proper_scores}, where the authors take advantage of \cite[Lemma 2.2]{new_multi_test2} or equivalently of \cite[Identity (17)]{new_multi_tes1}.
\\
\begin{lemma}
    Let $X_1$, $X_2$, $Y_1$, $Y_2$ be independent real random variables with finite expectations. Let $X_1,X_2$ be identically distributed with distribution function $F$ and let $Y_1,Y_2$ be identically distributed with distribution function $G$. Then
   \begin{equation}\label{eq:crps_expectation}
    \mathbb{E}(|X_1-Y_1|)-\frac{1}{2}\mathbb{E}(|X_1-X_2|)-\frac{1}{2}\mathbb{E}(|Y_1-Y_2|)=\int\limits_{-\infty}^{\infty}\left(F(x)-G(x)\right)^2dx
\end{equation}
\end{lemma}
Notice that, in our case, Equation \ref{eq:crps_expectation}, the distribution $G$ of $Y_1$ and $Y_2$ is degenerate, with all probability mass on a single point $y$.
%  (x notation will need to be cleaned here). Since $G(t)=:$
It follows that the third addend in the summation is zero. That is because $Y_1$ and $Y_2$ both following distribution $G$ implies that $\mathbb{E}(|Y_1-Y_2|)$ corresponds to the difference of two equal constant numbers.

### 2024-06-05 thesis/input_files/paper1.tex
cleaned all problems and warnings
final thesis

While $k([2,4], [8,9])=(16+36)^2=2704$
We can see that the latter calculation is much quicker and compact.
This gives a taste of how powerful kernels are thanks to the kernel trick.
Indeed, the idea of the the kernel trick can be extended to feature maps
$\phi$ involving an infinite feature space. In these cases, calculating the
respective kernel is equal to calculating the inner product between an infinite
number of features of the data points.

@@

\begin{definition}
    A sequence $\{v_n\}_{n=1}^{\infty}$ of elements of a normed vector space $\mathcal{V}$ is a Cauchy sequence if for every $\epsilon>0$, there exist $N=N(\epsilon) \in \mathbb{N}$ such that $\|v_n-v_m\|_{\mathcal{V}}<\epsilon \ \ \forall m,n\geq N$  
\end{definition}

@@

    $$  
    \begin{aligned}
    z^\intercal (K_1+K_2) z&= z^\intercal K_1 z + z^\intercal K_2 z \geq 0
    \\
    z^\intercal aK z&= a z^\intercal K z \geq 0
    \end{aligned}
    $$
    Let $K=K_1 \otimes K_2$, we have that the eigenvalues of $K$ are made up by the product pairs of the eigenvalues of $K_1$ and $K_2$. Thus, $K$ is positive semidefinite. Next, the matrix given by $k_1 \cdot k_2$ corresponds to the Schur product of $K_1$ and $K_2$, denote it $H$. Notice, $H$ is a principal submatrix of $K$, hence for any $z \in \mathbb{R}^n$ there exist a $t \in \mathbb{R}^{n^2}$ such that $z^\intercal H z= t^\intercal K t$. Using the fact that K is positive semidefinite we have that for any $z$, it holds 
    $$
    \begin{aligned}
        z^\intercal H z= t^\intercal K t\geq 0
    \end{aligned}
    $$
\end{proof}

\section{Kernel families}\label{kernel families}
The selection of the kernel determines the class of functions that will be searched by the learning algorithm. Prior knowledge of the problem domain can help restricting the candidate families of kernel functions.
Table \ref{tab:kernel types} contains popular kernel families in literature and applications.
\begin{table}
    \caption{Kernel types}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{lll}
        \toprule
       Kernel function & Equation & Hyperparameters \\
       \midrule
       Linear &  $k(x,y)=x^\intercal y$ &   \\
       Polynomial &  $k(x,y)=(x^\intercal y+c)^d$ &   c, d\\
       Gaussian RBF &  $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}^2}{2 l^2}\right)$ &   $l$\\
       Exponential RBF/Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}}{l}\right)$ &   $l$\\
       Absolute Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{1}}{l}\right)$ &   $l$\\
       Hyperbolic/Sigmoid Kernel &  $k(x,y2)=\tanh(\gamma x^\intercal y+r)$ &  $\gamma$, r \\
       Periodic &  $k(x,y)=\exp\left(\frac{-2 sin^2\left(\frac{\pi}{p}\|x-y\|_{2} \right)}{l^2}\right) $ &   $p, l$\\
       Chi-squared kernel &  $k(x,y)=\exp\left(-\gamma \sum\limits_i \frac{(x_i - y_i)^2}{x_i+y_i}\right) $ &   $\gamma$\\
        Cosine  &  $k(x,y)=\frac{x^\intercal y}{\|x\|_{2}\|y\|_{2}}$ &   \\
       Matern  &  $k(x, y) =  \frac{1}{\Gamma(\nu)2^{\nu-1}}\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}
        \Bigg)^\nu Y_\nu\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}\Bigg)$ &  $l, \nu$ \\

@@

% Any algorithm involving inner products between inputs can be combined with a kernel method. The chosen kernel function will calculate the inner product between the images of two inputs in the feature space. This makes the orginal algorithm operate in a high dimesional space. 
Kernel methods show good modularity because they work the same for any kernel and any data type. This concept is illustrated in Figure \ref{fig:workflow_kernels}, where an abstract kernel methods workflow is visualized. First we build the kernel matrix by using a kernel function to process the input data. Next, the kernel matrix is passed to the pattern analysis (PA) algorithm, which returns the learned function.

>>>>

while $$k([2,4], [8,9])=(16+36)^2=2704$$

We can see that the latter calculation is much quicker and compact.
This gives a taste of how powerful kernels are thanks to the kernel trick.
Indeed, the idea of the the kernel trick can be extended to feature maps
$\phi$ involving an infinite feature space. In these cases, calculating the
respective kernel is equal to calculating the inner product between an infinite
number of features of the data points.

@@

\begin{definition}
    A sequence $\{v_n\}_{n=1}^{\infty}$ of elements of a normed vector space $\mathcal{V}$ is a Cauchy sequence if for every $\epsilon>0$, there exists $N=N(\epsilon) \in \mathbb{N}$ such that $\|v_n-v_m\|_{\mathcal{V}}<\epsilon \ \ \forall m,n\geq N$  
\end{definition}

@@

    $$  
    \begin{aligned}
    z^\intercal (K_1+K_2) z&= z^\intercal K_1 z + z^\intercal K_2 z \geq 0
    \\
    z^\intercal aK z&= a z^\intercal K z \geq 0
    \end{aligned}
    $$
    Let $K=K_1 \otimes K_2$, we have that the eigenvalues of $K$ are made up by the product pairs of the eigenvalues of $K_1$ and $K_2$. Thus, $K$ is positive semidefinite. Next, the matrix given by $k_1 \cdot k_2$ corresponds to the Schur product of $K_1$ and $K_2$, denote it $H$. Notice, $H$ is a principal submatrix of $K$, hence for any $z \in \mathbb{R}^n$ there exists a $t \in \mathbb{R}^{n^2}$ such that $z^\intercal H z= t^\intercal K t$. Using the fact that K is positive semidefinite we have that for any $z$, it holds 
    $$
    \begin{aligned}
        z^\intercal H z= t^\intercal K t\geq 0
    \end{aligned}
    $$
\end{proof}

\section{Kernel families}\label{kernel families}
The selection of the kernel determines the class of functions that will be searched by the learning algorithm. Prior knowledge of the problem domain can help restricting the candidate families of kernel functions.
Table \ref{tab:kernel types} contains popular kernel families in literature and applications.
\begin{table}
    \caption{Kernel types}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{lll}
        \toprule
       Kernel function & Equation & Hyperparameters \\
       \midrule
       Linear &  $k(x,y)=x^\intercal y$ &   \\
       Polynomial &  $k(x,y)=(x^\intercal y+c)^d$ &   c, d\\
       Gaussian RBF &  $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}^2}{2 l^2}\right)$ &   $l$\\
       Exponential RBF/Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}}{l}\right)$ &   $l$\\
       Absolute Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{1}}{l}\right)$ &   $l$\\
       Hyperbolic/Sigmoid &  $k(x,y)=\tanh(\gamma x^\intercal y+r)$ &  $\gamma$, r \\
       Periodic &  $k(x,y)=\exp\left(\frac{-2 sin^2\left(\frac{\pi}{p}\|x-y\|_{2} \right)}{l^2}\right) $ &   $p, l$\\
       Chi-squared &  $k(x,y)=\exp\left(-\gamma \sum\limits_i \frac{(x_i - y_i)^2}{x_i+y_i}\right) $ &   $\gamma$\\
        Cosine  &  $k(x,y)=\frac{x^\intercal y}{\|x\|_{2}\|y\|_{2}}$ &   \\
       Matern  &  $k(x, y) =  \frac{1}{\Gamma(\nu)2^{\nu-1}}\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}
        \Bigg)^\nu Y_\nu\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}\Bigg)$ &  $l, \nu$ \\

@@

% Any algorithm involving inner products between inputs can be combined with a kernel method. The chosen kernel function will calculate the inner product between the images of two inputs in the feature space. This makes the orginal algorithm operate in a high dimesional space. 
Kernel methods show good modularity because they work the same for any kernel and any data type. This concept is illustrated in Figure \ref{fig:workflow_kernels}, where an abstract kernel methods workflow is visualised. First we build the kernel matrix by using a kernel function to process the input data. Next, the kernel matrix is passed to the pattern analysis (PA) algorithm, which returns the learned function.
